Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-12 doc/manuscript/ESSDmanuscript.Rmd
work on IPCC methods & definitions section

To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
In the case of forests, emission factors include carbon stocks and increments within various pools [@ipcc_2019_2019].

@@

Here, we
(1) review IPCC definitions of relevant carbon stocks and increments;
(2) describe mapping of ForC to IPCC's EFDB;
(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data transfer to EFDB;
(4) summarize the data in ForC that's relevant to EFDB and records that have been transferred to date; and
(5) provide recommendations for improving data collection, analysis, database, and accounting.

# IPCC definitions of carbon stocks and incremenets

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly for the purpose of national inventories [@ipcc_2006_2006;@ipcc_2019_2019]. 
More proximately, IPCC guidelines focus on:
(1) C stocks -- i.e., stores of C that would be vulnerable to release to the atmosphere upon land use change
and (2) C increments -- i.e., changes in those C stocks over one or more years.
*[detail how these are used]*

## Carbon stocks
Forest ecosystem C stocks may be parsed into pools in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies. 
IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
Quantifying these requires a one-time measurement.

@@

@@

@@

## Carbon increments

C increments are defined as the change over time, in annual increments, in each C pool.
These may be estimated as the difference between C stocks at two time points, or as the difference between inputs and outputs to the pool (i.e., fluxes).
Quantifying these requires at least two measurements.
(*But, Can carbon increments be inferred from a single measure and a known age (i.e., the approach we used in GROA)?*)
Fluxes are the inputs and outputs to each pool.

\clearpage
```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH). Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
knitr::include_graphics("figures_tables/C_variable_mapping.png")
```

@@

Mapping of variables is shown in Fig. `r `fig_variable_mapping`.

@@

>>>>

To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
In the case of forests, emission factors include carbon stocks, increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].

@@

Here, we
(1) review IPCC methods and definitions for tallying forest C;
(2) describe mapping of ForC to IPCC's EFDB;
(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data transfer to EFDB;
(4) summarize the data in ForC that's relevant to EFDB and records that have been transferred to date; and
(5) provide recommendations for improving data collection, analysis, database, and accounting.

# IPCC methods and definitions

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019]. 
For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; Mg C yr^-1^) are calculated as the sum of changes in various pools, including any harvested wood products.
IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig. `r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5-->[@ipcc_2006_2006]. 
Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools (Fig. `r fig_variable_mapping`).

@@

## Carbon pools

@@

Not required for forest land remaining forest land under Tier 1 methodology [@ipcc_2006_2006].

@@

```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH). Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
knitr::include_graphics("figures_tables/C_variable_mapping.png")
```

@@

Mapping of variables is shown in Fig. `r fig_variable_mapping`.

@@

Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies.

### 2022-04-12 doc/manuscript/ESSDmanuscript.tex
work on IPCC methods & definitions section

To improve the data accessible for C accounting, the IPCC created the
Emission Factor Database (EFDB;
\url{https://www.ipcc-nggip.iges.or.jp/EFDB/main.php}), which is
intended as a recognized library of emission factors and other
parameters that can be used for estimating greenhouse gas emissions and
removals. The EFDB can be used both for efforts to tally a nation's
intended or accomplished greenhouse gas reductions, or as a basis of
comparison for external parties to evaluate these inventories. The EFDB
encourages researchers to submit estimates of emission factors or other
related parameters from peer-reviewed journal papers or other accepted
sources for inclusion in the database. In the case of forests, emission
factors include carbon stocks and increments within various pools
\citep{ipcc_2019_2019}.

@@

Here, we (1) review IPCC definitions of relevant carbon stocks and
increments; (2) describe mapping of ForC to IPCC's EFDB; (3) describe
updates to ForC (ForC v4.0), most of which were implemented to
facilitate data transfer to EFDB; (4) summarize the data in ForC that's
relevant to EFDB and records that have been transferred to date; and (5)
provide recommendations for improving data collection, analysis,
database, and accounting.

\section{IPCC definitions of carbon stocks and incremenets}

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly for the purpose of national inventories
\citep{ipcc_2006_2006, ipcc_2019_2019}. More proximately, IPCC
guidelines focus on: (1) C stocks -- i.e., stores of C that would be
vulnerable to release to the atmosphere upon land use change and (2) C
increments -- i.e., changes in those C stocks over one or more years.
\emph{{[}detail how these are used{]}}

\subsection{Carbon stocks}

Forest ecosystem C stocks may be parsed into pools in various ways, and
while certain definitions and thresholds are more common than others,
there is no single standard for measuring or reporting that is adhered
to by all -- or even most -- studies. IPCC parses into biomass
(aboveground and belowground), dead organic matter (dead wood and
litter), and soil organic matter (Table 1). Quantifying these requires a
one-time measurement.

@@

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{major sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory\\
\hline
belowground biomass & all biomass of live roots & allometry, min dbh, assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
\hline
dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min dbh,  ... & default min dbh = 10cm, but may be chosen by country\\
\hline
litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & min dbh for dead wood , .. & includes entire O horizon\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & default sampling depth = 30cm, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

@@

@@

\subsection{Carbon increments}

C increments are defined as the change over time, in annual increments,
in each C pool. These may be estimated as the difference between C
stocks at two time points, or as the difference between inputs and
outputs to the pool (i.e., fluxes). Quantifying these requires at least
two measurements. (\emph{But, Can carbon increments be inferred from a
single measure and a known age (i.e., the approach we used in GROA)?})
Fluxes are the inputs and outputs to each pool.

\clearpage
\begin{figure}
\includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH). Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:fig_variable_mapping}
\end{figure}

@@

Mapping of variables is shown in Fig. \texttt{r}fig\_variable\_mapping`.

@@

As of April 11, 2022, ForC (v4.0) contained NA independent records (NA
total), NA of which were for variables relevant to EFDB (Fig. 1). These
records were distributed across all forested continents and ecozones,
with particularly high concentrations in \emph{{[}ecozone(s){]}} and low
concentrations in \emph{{[}ecozone(s){]}} (Fig. 2). The most widely
represented forest type was \emph{{[}type{]}}, followed by
\emph{{[}type{]}} and \emph{{[}type{]}} (Fig. 3).

@@

As of April 11, 2022, we had reviewed and sent NA, NA of which have been
reviewed, accepted, and posted (Fig. 2-3,
\texttt{Table\ r\ table\_variables}). \emph{{[}DETAILS{]}}

>>>>

To improve the data accessible for C accounting, the IPCC created the
Emission Factor Database (EFDB;
\url{https://www.ipcc-nggip.iges.or.jp/EFDB/main.php}), which is
intended as a recognized library of emission factors and other
parameters that can be used for estimating greenhouse gas emissions and
removals. The EFDB can be used both for efforts to tally a nation's
intended or accomplished greenhouse gas reductions, or as a basis of
comparison for external parties to evaluate these inventories. The EFDB
encourages researchers to submit estimates of emission factors or other
related parameters from peer-reviewed journal papers or other accepted
sources for inclusion in the database. In the case of forests, emission
factors include carbon stocks, increments (``stock changes''), and
fluxes (``gains'' and ``losses'') for various pools
\citep{ipcc_2006_2006, ipcc_2019_2019}.

@@

Here, we (1) review IPCC methods and definitions for tallying forest C;
(2) describe mapping of ForC to IPCC's EFDB; (3) describe updates to
ForC (ForC v4.0), most of which were implemented to facilitate data
transfer to EFDB; (4) summarize the data in ForC that's relevant to EFDB
and records that have been transferred to date; and (5) provide
recommendations for improving data collection, analysis, database, and
accounting.

\section{IPCC methods and definitions}

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly on a national level
\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
within a land-use category, annual stock changes (\(\Delta C\); Mg C
yr\textsuperscript{-1}) are calculated as the sum of changes in various
pools, including any harvested wood products. IPCC parses forest C pools
into biomass (aboveground and belowground), dead organic matter (dead
wood and litter), and soil organic matter (Table 1). For each pool,
\(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
takes the difference between gains and losses (influx and outflux
variables in Fig. 1), or using the ``Stock-Difference Method'', which
computes \(\Delta C\) based on C stocks at two points in
time\citep{ipcc_2006_2006}. Thus, C cycle variables relevant to the IPCC
methodology and to EFDB include C stocks, increments, and fluxes in the
IPCC-defined pools (Fig. 1).

@@

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation, both woody and herbaceous & min dbh & acceptable to exclude understory if minor component\\
\hline
 &  & biomass allometry & \\
\hline
 &  & include dead standing? & no\\
\hline
belowground biomass & all biomass of live roots & min dbh (coarse roots) & \\
\hline
 &  & allometry or assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & \\
\hline
 &  & min root diameter & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
\hline
 &  & sampling depth & \\
\hline
dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & include belowground? & \vphantom{1} yes\\
\hline
litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & layers included & includes entire O horizon\\
\hline
 &  & include belowground? & yes\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & depth = 30 cm default, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

\subsection{Carbon pools}

@@

Not required for forest land remaining forest land under Tier 1
methodology \citep{ipcc_2006_2006}.

@@

\begin{figure}
\includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH). Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:fig_variable_mapping}
\end{figure}

@@

Mapping of variables is shown in Fig. 1.

@@

As of April 12, 2022, ForC (v4.0) contained NA independent records (NA
total), NA of which were for variables relevant to EFDB (Fig. 1). These
records were distributed across all forested continents and ecozones,
with particularly high concentrations in \emph{{[}ecozone(s){]}} and low
concentrations in \emph{{[}ecozone(s){]}} (Fig. 2). The most widely
represented forest type was \emph{{[}type{]}}, followed by
\emph{{[}type{]}} and \emph{{[}type{]}} (Fig. 3).

@@

As of April 12, 2022, we had reviewed and sent NA, NA of which have been
reviewed, accepted, and posted (Fig. 2-3,
\texttt{Table\ r\ table\_variables}). \emph{{[}DETAILS{]}}
